# Statistics for Multiple Variables

Most of the time we want to look at the relationship between two or more variables.

R has several different ways of working with multiple variables. 


## Using formula notation

In R a "forumula" is created using the `~`  operator, which is found on the top left of the keyboard. 

In these examples the formula operator always works like this:
`dependent_variable ~ independent_variable` 
and if you have multiple independent_variables use a `+` to add them on the right.

One way to do crosstabs, also known as contingency tables, that uses the formula notation is to use the crosstab function
from lehmansociology.  This gives you the number of observations in each combination of the variables.

```{r}
lehmansociology::crosstab(cyl ~ gear, data = mtcars, title = "Cylinders by number of gears")
```

```{r}
lehmansociology::crosstab(cyl ~ gear, data = mtcars, title = "Cylinders by number of gears", format = "column_percent")
```


(You can also use the Janitor tabyl function to make cross tabs, this is explained in the next section.)


```{r message=FALSE}
library(Rmisc)
group.CI(weight ~ feed, 
         data = chickwts, ci = .90) 
```

You can also use the formula notation for a model with no independent variable by substituting a constant on
the right side.

```{r}
group.CI(weight ~ 1, 
         data = chickwts, ci = .90) 
```


### Ordinary Linear Model  
(This means it has an interval dependent variable.)
This works in the same way as group.CI, dependent variable on the left, independent on the right.

```{r}
lm(raises ~ critical, data = attitude)
```

### Generalized Linear Model
(In this case a logistic regression.)  
(in this code a dichtomous dependent variable is created using the cut() functions.)  

```{r}
USJudgeRatings$RTEN_d <- cut(USJudgeRatings$RTEN, median(USJudgeRatings$RTEN))

glm(RTEN_d ~ INTG, data = USJudgeRatings, family = binomial())
```
### Parametric t test
```{r}
t.test(extra ~ group, data = sleep)
```



## Using group_by 

Another way to look at the relationship between variables is to compare values of statistics for different groups.  

In this case one way to do this is with the group_by function from the `dplyr` package. 

Once you group your data there are a number of other functions within dplyr and in other packages that will use the 
groups to organize the results

```{r}
library(magrittr)
iris %>% dplyr::group_by(Species)  %>%
         dplyr::summarize(mean_sepal_length = mean(Sepal.Length),
                          median_sepal_legth = median(Sepal.Length),
                          Upper_CI = mean(Sepal.Length) + 1.96*sd(Sepal.Length),
                          Lower_CI = mean(Sepal.Length) - 1.96*sd(Sepal.Length)
                          )
```

```{r}
iris %>% dplyr::group_by(Species)  %>% skimr::skim()
```

## Entry of variables

Many R functions for multiple variables simply require that you enter the names of variables
in the correct order. Sometimes the names will need to be in quotation marks `"name"` and sometimes not.

### Cross tabs


To do a cross tab (also known as two-way table or contingency table) we also use `tabyl()`. The basic function
just gives counts for each combination. The first variable defines the rows and the second variable defines the
columns.

```{r}
library(magrittr)
library(janitor)
mtcars %>% tabyl(cyl, gear, show_na = FALSE)
```

Piping to `adorn_title() will add the name of the second variable.

```{r}
mtcars %>% tabyl(cyl, gear) %>%
           adorn_title()
```

To change to proportions use the `adorn_percentages()` functions.  In sociology we use the independent 
variable for the columns and the dependent variable for the rows. Therefore we use the column total for
the denominator `adorn_percentages(denominator = "col")`.  Sometimes you might want to calculate row or total
percents, in which case the denominator would be "row" or "all". 


```{r}
mtcars %>% tabyl(cyl, gear) %>%
      adorn_percentages(denominator = "col") %>%
      adorn_title()
```
To change the formatting to percentages pipe to the `adorn_pct_formatting()` function.


```{r}
mtcars %>% tabyl(cyl, gear) %>%
      adorn_percentages(denominator = "col") %>%
      adorn_pct_formatting() %>%
      adorn_ns() %>%
      adorn_title()
```

You can also display both the numbers and the percentages.

```{r}
mtcars %>% tabyl(cyl, gear) %>%
      adorn_percentages(denominator = "col") %>%
      adorn_pct_formatting() %>%
      adorn_ns() %>%
      adorn_title()
```
Adding a third variable will create separate tables for level of that variable.

```{r}
mtcars %>% tabyl(cyl, gear, am) %>%
      adorn_percentages(denominator = "col") %>%
      adorn_pct_formatting() %>%
      adorn_title()
```


You can also pipe to the `knitr::kable()` function to get a more polished table (see Chapter 2)


