# Statistics for Multiple Variables

Most of the time we want to look at the relationship between two or more variables.

R has several different ways of working with multiple variables. 

## Using formula notation

In R a "forumula" is created using the `~`  operator, which is found on the top left of the keyboard. 

In these examples the formula operator always works like this:
`dependent_variable ~ independent_variable` 
and if you have multiple independent_variables use a `+` to add them on the right.

*Crosstab using lehmansociology*
```{r}
lehmansociology::crosstab(tension ~ wool, data = warpbreaks)
```

*Ordinary Linear Model*  
(This means it has an interval dependent variable.)

```{r}
lm(raises ~ critical, data = attitude)
```

*Generalized Linear Model*
(In this case a logistic regression.)  
(in this code a dichtomous dependent variable is created using the cut() functions.)  

```{r}
USJudgeRatings$RTEN_d <- cut(USJudgeRatings$RTEN, median(USJudgeRatings$RTEN))

glm(RTEN_d ~ INTG, data = USJudgeRatings, family = binomial())
```

## Using summarize


## Other approaches

