# Graphs

In our classes we primarily use the ggplot2 package for graphing.

Remember htat you will need to load the package `library(ggplot2)`. 

As a package is is very flexible and provides many options that will let you do just about any visualization,
but this can make it seem complicated when you are starting out learning it.

## How ggplot works

Ggplot constructs a graph by building it up piece by piece, with the parts connected with `+` signs.

### ggplot()

A ggplot always starts with the `ggplot()` function.  For beginners, the important thing to know is that this
function sets up the axes for your plot.  To do this you n provide the name of the data set
and the name of the x variable and the y variable (if needed). Notice how the three axes below differ
depending on how many variables there are and what kind of variable they are.  The variable information
goes inside the `aes()` which is short for aesthetics.  

```{r}
library(ggplot2)

ggplot(iris, aes(x = Sepal.Length)) 

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width))

ggplot(iris, aes(x = Species, y = Sepal.Width))
```

### Adding geom_s
To add data to the axes you use a "geom" that determines the type of display.  The geom does not modify the
axis, it adds data to it. To show this, let's save the ggplot information and add to it.

Our first example is only for one variable.

```{r}
graph1 <- ggplot(iris, aes(x = Sepal.Length))  
graph1 + geom_histogram()
graph1 + geom_dotplot()
graph1 + geom_bar()
graph1 + geom_density()

```

Our second example uses two interval variables, so has different possibilities than the first. 
The second one shows one way that geoms can be combined.

```{r}
graph2 <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width))

graph2 + geom_point()
graph2 + geom_point() + geom_smooth()
```

This data includes two types of data.  The last example shows how multiple geoms can be combined.

```{r}
graph3 <- ggplot(iris, aes(x = Species, y = Sepal.Width))

graph3 + geom_point()
graph3 + geom_jitter()
graph3 + geom_boxplot()
graph3 + geom_boxplot() + geom_jitter()
```

### Add titles and labels
There are many additional options that you can use. This document will just show how to add a title
and labels.
```{r}
graph3 + geom_jitter() +
        ggtitle("Sepal Width by Species in the Iris Data") +
        labs(y="Sepal Width", x="Iris Species")  
```



 There are several websites that give detailed examples of more complex ways to create visualizations. 

## Common problems

Some of the most common problems in ggplot are the same as any other code in R, misspelling names, using 
the wrong graph for your data type, or having a problem with your tick marks. 

Here are some errors that you are most likely to see when using ggplot.

### unexpected symbol

This happens when you forget to put a `+` between the different parts of your plot.
```r
graph3  geom_jitter()
Error: unexpected symbol in "graph3  geom_jitter"
```

### invalid argument to unary operator

This happens if you have an extra `+`.

```r
 graph3 + + geom_jitter()
Error in +geom_jitter() : invalid argument to unary operator
```
### Error: unexpected ')' 

This happens if you have an extra closing parenthesis.

```r
graph2 <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)))
Error: unexpected ')' in "graph2 <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)))"

```


